---
title: "p8105_hw6_ns3782"
author: "NSK"
date: "2024-11-23"
output: github_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)

```

# Problem 1


```{r}

# Load weather data

```

# Problem 2: U.S. City Homicides

```{r}

# Load and clean homicide data

homicide_data <- 
  read_csv("homicide-data.csv") %>%
  mutate(
    city_state = paste(city, state, sep = ", "),
    solved = case_when(
      disposition == "Closed by arrest" ~ 1,
      disposition == "Closed without arrest" ~ 1,
      disposition == "Open/No arrest" ~ 0
    ),
    victim_age = as.numeric(victim_age)
  ) %>%
  filter(
    !(city_state %in% c("Dallas, TX", "Phoenix, AZ", "Kansas City, MO", "Tulsa, AL")),
    victim_race %in% c("White", "Black")
  )

```

## Adjusted odds ratios for solved homicides by victim sex in Baltimore, MD

```{r}

# Create new dataset filtering for Baltimore

baltimore_df <- 
  homicide_data %>%
  filter(city_state == "Baltimore, MD")

# Fit logistic regression

baltimore_solved_logistic <- 
  baltimore_df %>%
  glm(solved ~ victim_age + victim_race + victim_sex, 
      data = ., 
      family = binomial())

# Model results with estimate and confidence interval

baltimore_solved_logistic %>%
  broom::tidy(conf.int = TRUE) %>%
  mutate(
    OR = exp(estimate),
    OR_lower = exp(conf.low),
    OR_upper = exp(conf.high)
  ) %>%
  select(term, log_OR = estimate, OR, OR_lower, OR_upper, p.value) %>%
  knitr::kable(digits = 3)

```

Based on the logistic regression analysis of Baltimore homicides, the odds of solving a case are 0.35 times lower for male victims compared to female victims (OR = 0.35, 95% CI: 0.27, 0.47), after adjusting for victim age and race.

## Adjusted odds ratios for solved homicides by victim sex in 47 U.S. cities

```{r}

city_results <- homicide_data %>%
  group_by(city_state) %>%           
  nest() %>%                         
  mutate(
    results = map(data, ~glm(solved ~ victim_age + victim_race + victim_sex, 
                            data = ., family = binomial()) %>%
                    broom::tidy(conf.int = TRUE) %>%
                    filter(term == "victim_sexMale"))
  ) %>%
  unnest(results) %>%
  mutate(
    OR = exp(estimate),
    OR_lower = exp(conf.low),
    OR_upper = exp(conf.high)
  ) %>%
  select(city_state, OR, OR_lower, OR_upper, p.value)

```


```{r}

city_results %>%
  mutate(city_state = fct_reorder(city_state, OR)) %>%
  ggplot(aes(x = OR, y = city_state)) +
  geom_point() +
  geom_errorbarh(aes(xmin = OR_lower, xmax = OR_upper), height = 0.2) +
  labs(title = "Association Between Victim Sex and Solved Homicides by City",
       x = "Odds Ratio", 
       y = "City") 

```

